---
title: "Playing with penguins data"
author: "Paula Pappalardo"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: default
  word_document: default
---

## Setup

```{r setup, echo = T, warning = F, message = F, comment = ""}
library(knitr)

# tidy options

knitr::opts_chunk$set(echo = T, warning = F, message = F, comment = "")

# set up home directory

setwd("C:/Users/Paula/Dropbox/Smithsonian/Carpentries/penguins")
```

```{r libraries}
# install libraries
# install.packages("palmerpenguins")
# install.packages("magick")
# install.packages("visdat")

# load libraries
library(tidyverse)
library(palmerpenguins)
library(visdat)
library(cowplot)
library(magick)
```

## Get familiar with the data

Penguin artwork provided by @allison_horst. You can get it [here](https://allisonhorst.github.io/palmerpenguins/articles/art.html). The image I used below is called "penguins" and you need to save it in your home directory.

Penguins dataset: get a glimpse of the data and the distribution of missing values. You can find more information about this dataset [here](https://allisonhorst.github.io/palmerpenguins/articles/intro.html).

```{r data summaries}
# How many data points by species
n.byspp <- penguins %>%
  count(species)

# Averages for numeric variables
ave.penguins <- penguins %>%
  group_by(species) %>%
  summarize(across(where(is.numeric), mean, na.rm = TRUE))

# Take a glimpse at the data
glimpse(penguins)

# Where are the missing values? 
#visdat::vis_dat(penguins)
visdat::vis_miss(penguins)
```


## Make some plots

To learn how to add an image to a plot I found information [here](https://cran.r-project.org/web/packages/magick/vignettes/intro.html) and [here](https://wilkelab.org/cowplot/articles/drawing_with_on_plots.html).

&nbsp; 

First I customized theme parameters:

```{r plot setup}
niceplot <- theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
        axis.title.x = element_text(face = "bold", size = 14,
                                    margin = margin(t = 20, r = 0, b = 0, l = 0)),
        axis.title.y = element_text(face = "bold", size = 14,
                                    margin = margin(t = 0, r = 20, b = 0, l = 0)),
        axis.text.x = element_text(size = 13),
        axis.text.y = element_text(size = 13),
        legend.text = element_text(size = 13),
        legend.title = element_text(size = 13),
        legend.position = "bottom") 
```

Then I loaded the image and set up the plot:

```{r who is bigger, fig.width= 10, fig.height= 7}
# Load penguin art
fun.penguins <- image_read("penguins.png") %>%
  image_resize("570x380") %>%
  image_colorize(75, "white")


# Filter out NAs
mypenguins <- penguins %>% 
  dplyr::filter(!is.na(sex))

# Order species to match the image
mypenguins$f.species <- factor(mypenguins$species,
                               ordered = T,
                               levels = c("Chinstrap", "Gentoo", "Adelie"))
 
# plot body size by species for each sex
p <- ggplot(data = mypenguins, aes(x = f.species, y = body_mass_g, fill=sex)) +
  geom_boxplot(alpha = 0.4) +
  ggtitle("Who is bigger?") + 
  niceplot +
  labs(x = "Species", y = "Body size (g)") +
  theme(panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA))

# combine image and plot
ggdraw() + 
  draw_image(fun.penguins) + 
  draw_plot(p)
```

